---
title: "p8105_hw2_bh2849.Rmd"
author: "Binyue Hu"
date: "2022-10-03"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```

### Problem 1

Below we import and clean data from `NYC_Transit_Subway_Entrance_And_Exit_Data.csv`. The process begins with data import, updates variable names, and selects the columns that will be used in later parts fo this problem. We update `entry` from `yes` / `no` to a logical variable. As part of data import, we specify that `Route` columns 8-11 should be character for consistency with 1-7.

```{r}
trans_ent = 
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>% 
  janitor::clean_names() %>% 
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entry, exit_only, vending, entrance_type, 
    ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

As it stands, these data are not "tidy": route number should be a variable, as should route. That is, to obtain a tidy dataset we would need to convert `route` variables from wide to long format. This will be useful when focusing on specific routes, but may not be necessary when considering questions that focus on station-level variables. 

The following code chunk selects station name and line, and then uses `distinct()` to obtain all unique combinations. As a result, the number of rows in this dataset is the number of unique stations.

```{r}
trans_ent %>% 
  select(station_name, line) %>% 
  distinct
```

The next code chunk is similar, but filters according to ADA compliance as an initial step. This produces a dataframe in which the number of rows is the number of ADA compliant stations. 

```{r}
trans_ent %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

To compute the proportion of station entrances / exits without vending allow entrance, we first exclude station entrances that do not allow vending. Then, we focus on the `entry` variable -- this logical, so taking the mean will produce the desired proportion (recall that R will coerce logical to numeric in cases like this).

```{r}
trans_ent %>% 
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean
```

Lastly, we write a code chunk to identify stations that serve the A train, and to assess how many of these are ADA compliant. As a first step, we tidy the data as alluded to previously; that is, we convert `route` from wide to long format. After this step, we can use tools from previous parts of the question (filtering to focus on the A train, and on ADA compliance; selecting and using `distinct` to obtain dataframes with the required stations in rows).

```{r}
trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>% 
  select(station_name, line) %>% 
  distinct

trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```


### Problem 2

#### Read and clean the Mr. Trash Wheel sheet
```{r}
mrtrash_data = 
  read_excel("./data/Trash Wheel Collection Data.xlsx",
            sheet = "Mr. Trash Wheel",
            range = "A2:N549") %>%
  janitor::clean_names() %>%
  drop_na(c(dumpster)) %>%
  mutate(sports_balls = as.integer(round(sports_balls)),
         dataset = "Mr")
```


#### Read and clean Professor Trash Wheel sheet 
```{r}
protrash_data = 
  read_excel("./data/Trash Wheel Collection Data.xlsx",
             sheet = "Professor Trash Wheel",
             range = "A2:M96") %>%
  janitor::clean_names() %>%
  drop_na(c(dumpster)) %>%
  mutate(dataset = "Pro")
  
```

#### Combine Two Sheets

```{r}
mrtrash_data$year = as.numeric(mrtrash_data$year)
protrash_data$year = as.numeric(protrash_data$year)

trash_tidy = 
  full_join(mrtrash_data,protrash_data )
```

#### About These Data

* `trash-tidy` dataset is the combination of two excel sheets, there are 15 key variables, for example month, date, weight_tons, and glass_bottles. 
* The data frame has **`r nrow(trash_tidy)`** rows and **`r ncol(trash_tidy)`** columns. 
* The total weight of trash collected by Professor Trash Wheel is **`r sum(protrash_data$weight_tons)`** tons.
* The total number of sports balls collected by Mr. Trash Wheel in 2020 is **`r mrtrash_data %>% filter(year == 2020) %>% pull(sports_balls) %>% sum`**.
 

### Problem 3

#### Clean Data in pols-month.csv

```{r}
pols_data = 
  read.csv("./Data/fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(mon, c("year","month","day")) %>%
  mutate(year = as.integer(year),
         day = as.integer(day),
         month = as.integer(month),
         month = month.name[month],
         president = case_when(prez_dem == "1" ~ "dem", 
                               prez_gop == "1" ~ "gop")) %>% 
  select(-day, -prez_dem, -prez_gop)

```

#### Clean Data in snp.csv
```{r}
snp_data = 
  read_csv("./Data/fivethirtyeight_datasets/snp.csv") %>% 
  janitor::clean_names() %>% 
  mutate(date = as.Date(date, "%m/%d/%y")) %>% 
  separate(date, c("year","month", "day")) %>%
  mutate(year = as.integer(year),
         day = as.integer(day),
         month = as.integer(month),
         month = month.name[month]) %>%
  arrange(year, month)

```

### Tidy the Unemployment Data 

```{r}
unemp_data = 
  read_csv("./Data/fivethirtyeight_datasets/unemployment.csv") %>%
  janitor::clean_names() %>% 
  pivot_longer(jan:dec,
               names_to = "month",
               values_to = "percentage of unemployment") %>%
  mutate(year = as.integer(year),
         month = recode(month,
                   jan = "Janurary",
                   feb = "Feburary", 
                   mar = "March", 
                   apr = "April",
                   may = "May",
                   jun = "June",
                   jul = "July",
                   aug = "August",
                   sep = "September",
                   oct = "October",
                   nov = "November",
                   dec = "December")) 

```

#### Join the Datasets 

```{r}
join_data = 
  left_join(pols_data, snp_data, by = c("year", "month")) %>%
  left_join(unemp_data, by = c("year", "month"))
```

#### About These Datasets 

* The `pols_data` dataset includes `r count(pols_data)` observations and `r ncol(pols_data)` variables, the following variables: `r names(pols_data)`. 
* The `snp_data` dataset includes `r count(snp_data)` observations and `r ncol(snp_data)` variables, the following variables: `r names(snp_data)`.
* Tthe `unemp_data` dataset includes `r count(unemp_data)` observations and `r ncol(unemp_data)` variables, the following variables: `r names(unemp_data)`.

* The joined dataset, `join_data`, includes `r count(join_data)` observations and `r ncol(join_data)` variables. 
The dimension of the joined dataset is `r dim(join_data)`. 
The range of years is between `r min(join_data$year)` and `r max(join_data$year)`. 
There are key variables in the dataset like `president`, `unemployment_rate`, and `close`. 


